### 2016-07-06 src/5_2_Proposal.tex
change order chapters

\chapter{Adaptive parallel algorithm for Vector error correction model}

@@

Vector error correction model (VECM) parameters are obtained using ordinary
least squares (OLS) method \cite{golub1980}, which is computationally expensive
when the number of past values $L$ and observations $p$ increases. Moreover,
obtaining cointegration vectors is also an expensive routine because of the use
of the Johansen method.

In this study we found that for some combinations of $L$ and $p$, the number of
cointegration relations is increased and this leads to better performance. We
propose to parallelise this search in order to achive faster execution times.

\section{The problem} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the most recent historical data. However, one of the main limitations of this
methodology is that it assumes cointegration vectors do not change in time.  In
fact, the long-run relationships between the time series might change due to
several economic factors that can lead to structural breaks in the cointegrating
relation \cite{gregoryETal1996}.

In this study, we show that cointegration relations depend on the amount $L$ of
historical data considered, and how many lags $p$ were used. This search is
computationally expensive, considering that the Johansen
method~\cite{johansen1995} (which is of order $O(n^3)$) is used and these
parameters change at every iteration.

Our proposal consists on to parallelise this search in order to get new
parameters at every iteration.

In finance, little academic reasearch has been done using high-frequency Forex
rates \cite{Bekiros2015}. In our study, tests were run using four currency
rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY)
(USDJPY) with a 10-seconds frequency.  The forecast capability of our method was
later measured using MAPE, MAE and RMSE \cite{armstrong1992}.

\section{Methodology}

VECM is obtained re-writing the VAR model, shown in equation (\ref{eq:var}), in
terms of differences defines as: $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}

If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{B}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation.

\subsection{Motivation}
The relation between the number of cointegration relations
and parameters $L$ and $p$ is studied and then a relation between cointegration
and performance is stated.

Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
the Johansen method for different values of the parameters $L$ leaving $p$ set
as $1$, using four currencies data, i.e., $l=4$, and 1000 considering iterations
($it=1000$).
\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/histCointVectors}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (100, 1900, 6100 and 10700).
  In the cases of $L=100$ and $L=6100$, most of the time there is little
  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
  $L=10700$ significant cointegration is found.}
  \label{fig:hists}
\end{figure}
We know that when no cointegration vectors can be found, i.e $r=0$, implies that
the variables are not cointegrated. In the same way, when $r=l$ reveals that
none of the processes is I(1) but, they are all stationary.  The interesting cases of
cointegration are those, where $r$ lies strictly between $0$ and $l$, i.e.
$0<r<l$,

Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
which corresponds to the opening of the New York financial market.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
  \caption{Histogram of the number of cointegration vectors using $p=5$. 
   Despite the change in the distribution of cointegration vectors, 
   the percentage of cointegration is nearly the same as in the case $p=1$. 
   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
   vectors is very similar to 1 cointegration vector for the case $L=10700$
   in figure \ref{fig:hists}.}
   \label{fig:histsp5}
\end{figure}

The same experiments, setting $p=5$, were carried out in order to see the
effect on the number of cointegration vectors found. 
Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
of the number of cointegration vectors, the sum of the cases where $0<r<4$,
remains nearly the same. 
Therefore, we can say that the number of lags doesn't significantly affect
the extent of cointegration found.

In order to measure the extent of cointegration, we introduce a
{\em percentage of cointegration\/} as following:
\begin{equation} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation}
where c.v. stands for cointegration vectors and $it$ is the number of iterations.

The goal of our next experiment was to find a relation between this ratio
$PC$ and the performance of the accuracy measure MAPE (see equation 
\ref{eq:MAPE}). 
The experiments suggest that there is indeed a relation between the $PC$ ratio
and MAPE.
Figure~\ref{fig:cointvsmape} shows that MAPE rapidly decreases with increasing
$PC$.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.7\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
  Best $L$ found was 8900.}
  \label{fig:cointvsmape}
\end{figure}

From these results we observe that cointegration changes with different values
for $L$ and $p$ and also that better cointegration percentage leads to better
accuracy performance.  Therefore we propose to choose best $L$ and $p$
parameters determined at every step.  The criterion for choosing $L$ and $p$ is
to maximise the percentage of cointegration $PC$ considering a number of past
iterations $it$.

Algorithm ~\ref{alg:AVECM} solves equation \ref{eq:vareq} defining first
parameters $L$ and $p$ using the function {\bf get\_best\_params} which makes a
grid search on the two vector lists $Ls$ and $ps$ and returns the parameters $L$
and $p$ which maximise the percentage of cointegration $PC$ (see
equation~\ref{eq:pcoint}) for a pre-defined number of iterations.  Since this
search is computational expensive, the number of cointegration vectors at each
iteration is computed in parallel, thus ensuring a response before new data is
available.  This parallel routine also allows to refine our grid search leading
to obtain improved forecasting performance.  After that optimum $L$ and $p$
parameters are found, VECM is built and used to forecast next data.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$ps$: vector with the number of past values range \\
$Ls$: vector with window sizes range ($L<N$) \\
$outit$: Number of out-of-sample iterations($outit < N-L$)
$it$: Number of iterations of testing (in-sample)\\
$of$: Starting point of Testing \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[1],\dots,\mathbf{y}_{\text{pred}}[N-L]\}$: model predictions 
\FOR { $i =of$ to $of + outit$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[i-\texttt{max}(Ls):i]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,it,Y)$
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i-L:i]$
        \STATE $model = VECM(\mathbf{y}_i, p)$
        \STATE $\mathbf{y}_{\text{pred}}[i] = model.predict()$
\ENDFOR
\end{algorithmic}
\caption{AVECM: Adaptive VECM.}
\label{alg:AVECM}
\end{algorithm}

\section{Experimental results}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from the free
database Dukascopy which gives access to the Swiss Foreign Exchange marketplace
~\cite{Dukascopy2014}.

The tests were done using 10-seconds frequency from ask prices which
corresponded to 8640 data points per day from the 11th to the 15th of August
2014. Since data datetimes were in GMT, opening market times for London, New
York, Sidney and Tokyo corresponded with data points 1440, 3240, 6120 and 6840
respectively. Our tests were made considering two hours after the opening time
of the 13th of August. Therefore, in order to obtain these times, we added 2
days and 2 hours to all the opening times ($2\times 8640 + 2 \times 360$) to obtain:
19440, 21240, 24120 and 24840. We called offset to these times in
table~\ref{tab:stats}, which allowed us to obtain performance at different
times of the day.

\subsection{Unit root tests} \label{sec:unitroot}
Before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test at 95\% significance level.
Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore AVECM.

\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          &  -0.05398   & -1.94101 & True       \\
$\Delta$ EURUSD & -89.12344   & -1.94101 & False       \\
GBPUSD          &  -0.88208   & -1.94101 & True          \\
$\Delta$ GBPUSD & -82.56501   & -1.94101 & False       \\
CHFUSD          & -0.49941    & -1.94101 & True         \\
$\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
JPYUSD          &  0.36721    & -1.94101 & True        \\
$\Delta$ JPYUSD & -101.15589  & -1.94101 & False     \\ 
\hline
\end{tabular}
\end{center}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
frequency.}
\label{tab:adf}
\end{table}

\section{Parallel implementation} \label{sec:paralell}

To determine optimal $L$ and $p$ parameters require the using of the Johansen
method which is a computationally expensive routine. 
In order to improve the execution time of this search, our proposal included
a parallel search of VECM parameters using high performance computing.
The main objective is to obtain a response before a new data arrives in
the next 10 seconds.

The Johansen method is implemented in the Python Statsmodels
library~\cite{seabold2010} and the parallel implementation was done using
MPI in Python. 
MPI was chosen because it allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
Our tests were done in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.

Since financial market are constantly changing, we ran our
algorithm using different times of the day. We chose to use the algorithm
starting on the third day and two hours after the market opening times. These
times corresponded to data points 19440 (5 GMT London), 21240 (10 GMT New
York), 24120 (19 GMT Sydney) and 24840 (9 GMT Tokyo).

In order to determine the percentage of cointegration vectors, we considered
different number of iterations: 10, 50 and 100. The number of possible
combinations ($nparam$) of $L$ and $p$ is shown for 12, 24 and 47.

@@

Table~\ref{tab:stats} shows the out-of-sample performance measures: MAPE, MAE
and RMSE for AVECM for different offsets, iterations and number of parameters.
We observed that performance measures differ for different offsets having even
in some cases different orders. London and New York offsets are order $10^{-3}$
while Sidney and Tokyo vary between $10^{-3}$ and $10^{-4}$. Despite this
difference in the order, they all achieve their minimum MAPE, MAE and RMSE
measurements at 100 iterations. In the case of Tokyo, it also reduces its MAPE
order. In all cases, we can see that performance measures are improving with
the number of parameters.  This is because there are more possible options to find
optimal cointegration parameters.

>>>>

\chapter{Online VECM proposal}
Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated. The Vector Error Correction
Model (VECM) is an econometric model which characterises the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium. In this study, we propose an Online VEC model (OVECM) which
optimises how model parameters are obtained using a sliding window of the most
recent data. Our proposal also takes advantage of the long-run relationship
between the time series in order to obtain improved execution times. Our
proposed method is tested using four foreign exchange rates with a frequency of
1-minute, all related to the USD currency base. OVECM is compared with VECM and
ARIMA models in terms of forecasting accuracy and execution times. We show that
OVECM outperforms ARIMA forecasting and enables execution time to
be reduced considerably while maintaining good accuracy levels compared with
VECM.

@@

\section{The problem}
In finance, it is common to find variables with a long-run and/or a short-run
equilibrium relationship. This relationship is called cointegration and it
reflects the idea of that some set of variables cannot wander too far away from
each other.

Cointegration means that one or more combinations of these variables is
stationary even though individually they are not.

Some financial models, such as  vector error correction model (VECM), take advantage of
this property and describe the joint behaviour of several cointegrated
financial instruments.

VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
model which express future values in terms of its own history and other
variables past values. If cointegration exists, VECM allows to introduce an
error correction term due to its cointegration estimation error.

VECM as well as VAR model parameters are obtained using ordinary least squares
method (OLS). Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations considered increases. Moreover, OLS is an ill-posed problem which
admits an infinite number of solutions.

Ridge regression method (RR) tackles this ill-posed problem and is usually
formulated instead of OLS. RR includes a regularisation parameter that leads to
better generalisation capability. However, RR is still computationally
expensive dealing with large datasets. Recently, online learning algorithms
have been proposed to solve problems with large data sets because of their
simplicity and their ability of updating the model when new data is available.

We propose an online formulation of the VEC model based on the
AAR method considering only a sliding window of the most recent data. The
algorithm introduces matrix optimisations in order to reduce the number of
operations. Our method is later tested with financial data from the foreign
exchange market.

\section{Methodology} \label{sec:methodology}

Since VECM parameter estimation is computationally expensive, we propose an
online version of VECM (OVECM).  OVECM considers only a sliding window of the
most recent data. Moreover, since cointegration vectors represent long-run
relationships which vary little in time, OVECM determines firstly if they require calculation.

OVECM also implements matrix optimisations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.

Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
following:

\begin{itemize}
\item The function \texttt{getJohansen} returns cointegration vectors given by
the Johansen method considering the trace statistic test at 95\%
significance level.
\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
and~(\ref{eq:vecY}) that allows VECM to be solved.
\item The function \texttt{vecMatrixOnline} returns the
matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
\item Out-of-sample outputs are saved in the variables 
$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
\item The model is solved using OLS.
\item In-sample outputs are saved in the variables $\Delta
\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
series.
\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
MAPEs are stored in the variable $e$. If the average of MAPEs
($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
\item If new cointegration vectors are required, the function
\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
\end{itemize}

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
$\text{mean\_error}$: MAPE threshold \\
$n$: interpolation points to obtain MAPE \\
\ENSURE  $\,$ \\
$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
	\IF {i = 0 }
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
	\ELSE
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
    \ENDIF
    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$\mathbf{A} \gets
        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \ENDIF
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
\STATE $\mathbf{Y}_{\text{pred}}\gets 
\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
\end{algorithmic}
\caption{OVECM: Online VECM}
\label{alg:proposal}
\end{algorithm}

A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.

\resizebox{\textwidth}{!}{
   \begin{tikzpicture}[node distance=2cm]
    \label{fig:pseudocodigo}
    %\node (start) [startstop] {Start};
    %\node (in1) [io, below of=start] {Input};
    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};

    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};

    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };

    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};

    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $};

    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };

    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};

    %\node (stop) [startstop, below of=out1] {Stop};

    %\draw [arrow] (start) -- (in1);
    \draw [arrow] (in1) -- (pro1);
    \draw [arrow] (pro1) -- (dec1);
    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
    \draw [arrow] (pro2a) -- (pro3);
    \draw [arrow] (pro2b) |- (pro3);
    \draw [arrow] (pro3) -- (dec2);
    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
    \draw [arrow] (pro4a) |- (dec3);

    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
    \draw [arrow] (pro5) |- (pro1);
    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
    %\draw [arrow] (out1) -- (stop);
    \end{tikzpicture}
}

Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
to an online context in order to get a reasonable comparison with our proposal
(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
called with a sliding window of the most recent data, whereby the models are
updated at every time step.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
        \STATE $model = VECM(\mathbf{y}_i, p)$
        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
\end{algorithmic}
\caption{SLVECM: Sliding window VECM}
\label{alg:SLVECM}
\end{algorithm}

Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
executed for every time series.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: autoregressive order \\
$d$: integrated order\\
$q$: moving average order\\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
\FOR { $j =0$ to $l-1$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
\ENDFOR
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
\end{algorithmic}
\caption{SLARIMA: Sliding window ARIMA}
\label{alg:SLARIMA}
\end{algorithm}

Both OVECM and SLVECM time complexity is dominated by Johansen method which is
$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
iterations.

\section{Results} \label{sec:results}

\subsection{Data} \label{sec:unitroot}
Tests of SLVECM, SLARIMA and our proposal OVECM were carried out using foreign
four exchange data rates: EURUSD, GBPUSD, USDCHF and USDJPY. This data was
collected from the free database Dukascopy which gives access to the Swiss
Foreign Exchange Marketplace ~\cite{Dukascopy2014}.

The reciprocal of the last two rates (CHFUSD, JPYUSD) were used in order to
obtain the same base currency for all rates.  The tests were done using
1-minute frequency from ask prices which corresponded to 1.440 data points per
day from the 11th to the 15th of August 2014.

\subsection{Unit root tests} \label{sec:unitroot}
Before running the tests, we firstly checked if they were I(1) time series
using the Augmented Dickey Fuller (ADF) test.

\begin{table}[h!]
\caption{Unit roots tests}
\label{tab:adf}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          & -0.64 & -1.94 & True       \\
$\Delta$EURUSD & -70.45   & -1.94 & False       \\
GBPUSD          & -0.63   & -1.94 & True          \\
$\Delta$GBPUSD & -54.53   & -1.94 & False       \\
CHFUSD          & -0.88   & -1.94 & True         \\
$\Delta$CHFUSD & -98.98   & -1.94 & False       \\
JPYUSD          & -0.65 & -1.94 & True        \\
$\Delta$JPYUSD & -85.78 & -1.94 & False     \\ 
\hline
\end{tabular}
\end{center}
\end{table}

Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore OVECM.

\subsection{Parameter selection} \label{sec:paramselection}

In order to set OVECM parameters: windows size $L$ and lag order $p$,
we propose to use several window sizes: $L = 100, 400, 700, 1000$. For every window size $L$ we chose lag order with minimum AIC.

ARIMA parameters were also obtained using AIC.
Parameters optimisation is presented in table~\ref{tab:params}:

\begin{table}[ht]
\caption{Parameters optimisation. VECM order and ARIMA parameters were selected
using AIC.}
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}

In OVECM we also define a mean\_error variable, which was defined based on the
in-sample MAPEs. Figure \ref{fig:mapes} shows how MAPE moves and how mean\_error
variable help us to decide whether new cointegration vectors are needed.

\begin{figure}[!h]
  \centering
   \includegraphics[width = 10.5cm]{img/mapes} 
  \caption{In-sample MAPEs example for 50 minutes. The average of them is
  considered to obtain new cointegration vectors.}
  \label{fig:mapes}
 \end{figure}

\subsection{Execution times} \label{sec:exectimes}
We ran OVECM and SLVECM 400 iterations. SLARIMA execution time is excluded
because its is not comparable with OVECM and SLVECM. SLARIMA was created based
on statsmodels library routine ARIMA.

The execution times are shown in the table~\ref{tab:extimes}.

\begin{table}[h!]
\caption{Execution times}
\label{tab:extimes}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& $L$ & order & e  & Time[s] \\
\hline
OVECM & 100 &p=2  & 0      & 2.492\\
OVECM & 100 &p=2  & 0.0026  & 1.606\\
SLVECM & 100 &p=2& -- & 2.100\\
\hline
OVECM & 400 & p=5  & 0      & 3.513\\
OVECM & 400 &p=5  & 0.0041  & 2.569\\
SLVECM & 400 & p=5 & -- & 3.222\\
\hline
OVECM & 700 &p=3  & 0      & 3.296\\
OVECM & 700 &p=3  & 0.0032  & 2.856\\
SLVECM & 700 &p=3 & -- & 3.581\\
\hline
OVECM & 1000 & p=3 & 0      & 4.387\\
OVECM & 1000 & p=3  & 0.0022  & 2.408\\
SLVECM & 1000 & p=3  & -- & 3.609\\
\hline
\end{tabular}
\end{center}
\end{table}

It is clear that execution time depends directly on $L$ and $p$ since they
determine the size of matrix $\mathbf{A}$ and therefore affect the OLS function
execution time.
It is worthy of note that execution time also depends on mean\_error because it
determines how many times OVECM will recalculate cointegration vectors which is
an expensive routine.

Figure~\ref{fig:mapes} shows an example of the in-sample MAPE for 50 iterations.
When the average of the in-sample MAPEs is above mean\_error new cointegration
vectors are obtained. In consequence, OVECM performance increases when
mean\_error increases. However, this could affect accuracy, but
table~\ref{tab:stats} shows that using an appropriate mean\_error doesn't affect
accuracy considerable.

@@

Table~\ref{tab:stats} shows in-sample and out-of-sample performance measures:
MAPE, MAE and RMSE for OVECM, SLVECM and SLARIMA. Test were done using the
parameters defined in table \ref{tab:params}.  We can see that OVECM has very
similar performance than SLVECM and this support the theory that cointegration
vectors vary little in time. Moreover, OVECM also out performed SLARIMA based on
these three performance measures.
